% =============================================================================
% 6. Business Analysis and Decision Framework -- target ~4.5 pp
% THE BA-CORE OF THE THESIS. This is what distinguishes a Business Analytics
% thesis from a generic ML report.
% Three sub-themes:
%   1. Why preference alignment is becoming a strategic capability
%   2. A decision framework for choosing a training method
%   3. The alignment economics that drive the framework
%
% NOTE ON SCOPE: this chapter must NOT re-narrate the technical history of
% RLHF, DPO, RAG or ICL. That work belongs in \cref{sec:bg}. By the time
% the reader reaches this chapter they have already met all four
% preference-optimisation losses and the ICL baseline. \S6.1 frames the
% market context; \S6.2 turns the results in \cref{sec:results} into a
% decision; \S6.3 prices that decision against current cost trajectories.
% =============================================================================

\section{Business Analysis and Decision Framework}
\label{sec:business}

\subsection{Market Tendencies}
\label{sec:business:tendencies}

The conversation about large language models has shifted over the last three years from "which model is the best" to "which model best fits this user." Frontier systems from OpenAI, Anthropic and Google sit within a few percentage points of each other on the public arenas \todo{cite LMSYS Arena}, and the practical decisions companies make about which one to deploy are increasingly driven by price, latency and integrations rather than raw capability. As performance commoditises at the base-model layer, the place where a product can still differentiate moves downstream of it, into how well the system fits the user in front of it. Personalisation is no longer a feature; it is the surface where competition now happens. The bottleneck on that surface is the start: a new user has given the product almost nothing to work with, and the choice of how to personalise from that handful of signals is what determines whether they bounce or stay.

The industry's first answer to this has been context. Rather than retrain a model for every user, products feed user-specific information back into the prompt at inference time and let the model condition on it. Claude Projects, ChatGPT Memory, and Gemini's long-context window \citep{reid2024gemini} are all expressions of the same bet: that a powerful enough frozen model can absorb a user's preferences from text alone. The bet has held up well for ad-hoc tasks but starts to strain when preferences accumulate over months and need to be reliably attended to, which is precisely the regime where training a small per-user adapter starts to look attractive. The two paradigms introduced in \cref{sec:bg} are not academic alternatives; they are the two strategies the market is actively trying out, and the question this thesis answers is when each one is the right call. The findings that follow come to inform organisations on how they can tackle these changes and leverage their data, or lack thereof, to enter the LLM world.

\subsection{Business Applications}
\label{sec:business:framework}

These findings become relevant in a business context mainly for companies that use LLMs for customer-facing sales and recommendation, and the natural way to see why is to think about what those companies already have. Almost every consumer business that has been operating for more than a few years sits on a large quantity of discrete user signal, the kind of data classical recommender systems were built to consume. Ratings, watch histories, purchase sequences, thumbs-ups and skips are all the same shape as the preference pairs used throughout this thesis. The work of \citet{bao2023tallrec} and \citet{friedman2023recllm} shows that this signal can be fed into an LLM either by training on it or by placing it in the prompt, and the survey by \citet{fan2024recllmsurvey} confirms this is the direction the field is moving. The question for a company is no longer whether to do it, but how.

That decision is what the results in \cref{sec:results} inform, and the answer is sharper than a single curve can show because the deployment format matters as much as the training size. The crossover in \cref{fig:primary_curves} tells a clean story when the product is going to score fixed options: at $n \leq 10$ ICL leads on both accuracy and cost; somewhere between $n = 10$ and $n = 50$ the trained methods catch the ICL plateau; and at $n = 100$ SimPO is clear of \texttt{icl\_chat} by 3.6 points on GoodReads and 5.9 points on Netflix, with the gap widening further at $n = 150$. A music app that ranks the next track from a known catalogue, a streaming service that orders a row of thumbnails, or a search engine that re-ranks ten retrieved candidates is operating in exactly the format the primary sweep measures. For these products the §5.1 ranking is the operational ranking, and the decision is the one the cold-start curves draw: stay in ICL until the average user accumulates enough preferences to cross the band, then switch to a trained adapter.

The result that complicates this picture is the free-form generation sub-study (\cref{tab:judge_results}). When the same models are asked to write a one-sentence recommendation rather than score a pair of fixed completions, the ranking inverts. \texttt{icl\_chat} at $n = 10$ wins the head-to-head judge comparison 65.5 percent of the time and names the user's held-out liked book 79.0 percent of the time ($p = 2.2 \times 10^{-9}$ against chance). SimPO at $n = 100$, the offline winner, drops to a 46 percent judge win rate and a 59 percent alignment rate, and writes recommendations that are statistically indistinguishable from the untrained base. The two name the same book on 94 of 100 cells. The fixed-completion gap between SimPO and the base model, which is forty percentage points on GoodReads, does not transfer to what the model actually says when asked to recommend.

This is a finding the AI community should sit with. A LoRA adapter of rank 16 trained for 100 epochs on 100 preference pairs visibly sharpens the log-probability margin between chosen and rejected completions in the exact format it was trained on, but it leaves the generation distribution of the underlying model essentially unchanged. The signal lands where the loss looks, and not anywhere else. The implication for the literature on preference optimisation is that pairwise accuracy on the training format is a partial proxy for personalisation in deployment, not a complete one, and benchmarks that report only the former may be over-stating the value of the methods they compare. For a product team the implication is more direct: the decision of which method to deploy depends not only on how many preferences each user has but on what the model will be asked to produce when serving them.

\begin{table}[h]
\centering
\caption{Decision regime by training size per user and by deployment format. The first column reproduces the regime split implied by \cref{tab:primary_acc}; the recommendation differs by row depending on whether the product scores fixed options or generates open text.}
\label{tab:business_regimes}
\begin{tabular}{lll}
\toprule
\textbf{Regime ($n$ per user)} & \textbf{Scoring / ranking product} & \textbf{Generative product} \\
\midrule
Cold-start floor ($n \leq 10$)        & ICL          & ICL \\
Transition band ($10 < n \leq 50$)    & Project engagement first   & ICL \\
Steady state ($n > 50$)               & SimPO (or IPO with anchor) & ICL, pending evidence \\
\bottomrule
\end{tabular}
\end{table}

Read together, the two results reshape the framework in \cref{tab:business_regimes}. For products whose user-facing surface is a ranked list or a structured choice the §5.1 ordering is the right guide and trained adapters are worth deploying once $n$ crosses fifty. For products whose user-facing surface is a sentence written by the model, which describes most of what the LLM era has actually shipped, ICL is the dominant strategy across every $n$ the experiment tested and the burden of proof for switching to a trained method should sit on the team proposing the switch rather than on the team defending the prompt. That is a much narrower mandate for per-user training than the primary curves alone would have suggested, and it is the kind of finding that only becomes visible when offline metrics are checked against the format the product actually uses.

What does not change between the two paths is the infrastructure. The signal a company needs to collect is the same paired or unpaired preference data either way, and the serving stack only diverges at the last step, with the trained methods loading a small LoRA adapter on top of the same frozen base model that ICL queries directly. \citet{sheng2024slora} showed that thousands of LoRA adapters can be served concurrently on shared infrastructure, which removes the operational objection that per-user training is impractical at scale. The broader point is that the LLM era does not retire the work companies have done on their recommendation stacks over the past decade, it relocates it. The years of clicks and ratings that fed a matrix-factorisation model do not become irrelevant when the front door turns into a chat box, they become the cold-start fuel for whatever personalisation method sits behind that chat box, and the same paired data can be reused under either paradigm if the team later decides to revisit the choice.

\subsection{Alignment Economics}
\label{sec:business:economics}

The decision of whether a company should use in-context learning or shift to a preference-optimisation method is not a one-time architectural choice. It depends on three variables that differ widely between products: how many preferences a user generates during onboarding, how often that same user queries the system afterwards, and what the model is asked to produce when it serves them. Mapping the experimental findings from \cref{sec:results} against the cost trajectories documented in the Stanford AI Index 2025 \citep{maslej2025aiindex} lets us turn the abstract trade-off into a decision a product team can actually make, and the format axis introduced in \cref{sec:business:framework} changes the arithmetic more than the headline curves would suggest.

Start with the case where the format favours training. For a scoring or ranking product at $n = 100$, the cost of running an inference at GPT-3.5 quality dropped from \$20.00 per million tokens in November 2022 to \$0.07 per million tokens by October 2024, a 280-fold reduction in roughly eighteen months \citep{maslej2025aiindex}. Hardware price-performance is improving at 30 percent per year, and a 3.8-billion-parameter Phi-3-mini now matches the MMLU score that a 540-billion-parameter PaLM achieved two years earlier. The practical effect is that the per-query cost of an ICL prompt, which used to be the dominant economic argument against the method, has collapsed. An ICL inference at $n = 100$ carries roughly four thousand extra prompt tokens compared to a trained-model inference. At the Stanford lower bound that adds less than three hundredths of a cent per query. At a more realistic mid-market rate of \$0.20 per million input tokens it is about eight hundredths of a cent. A LoRA training run for one user on a rented A100 costs somewhere between ten and twenty cents, which puts the break-even point at roughly two hundred and fifty queries per user without prompt caching, and around two thousand five hundred queries once prefix caching is in play.\footnote{Anthropic's prompt-caching documentation reports a roughly tenfold cost reduction on cached prefix tokens: \url{https://platform.claude.com/docs/en/build-with-claude/prompt-caching}.} For a scoring product these numbers favour training once a user is engaged enough to query the system a few hundred times.

The case where the format does not favour training looks very different. The break-even calculation above implicitly prices the value of a training run as the accuracy gap between SimPO and \texttt{icl\_chat} at $n = 100$, which is real for fixed-completion scoring. For a generative product that gap is not the right number to use. The free-form sub-study found that SimPO and the untrained base name the same recommendation on $94$ of $100$ cells and lose head-to-head against \texttt{icl\_chat} at $n = 10$ by 27 points on judge preference and 20 points on alignment. The marginal value of the training run, measured in what the user actually receives, is somewhere between negligible and negative. Spending ten to twenty cents per user to obtain a model that writes the same sentence as the base model and a worse sentence than the ICL prompt is not a break-even at any query volume, it is a loss at every query volume. The economic logic that argues for training when scoring is the same logic that argues against training when generating, because the assumed accuracy delta does not show up where the user can see it.

Two trajectory effects make this picture less stable than \cref{tab:business_regimes} suggests, and both lean further toward ICL. The first is that ICL economics are improving faster than training economics. \citet{maslej2025aiindex} reports inference cost falling between 9 and 900 times per year depending on the task, while hardware price-performance drops a steadier 30 percent per year. Over the next few cycles, ICL gets cheaper than training does, which pushes the break-even point higher even for the scoring case and rewards companies that delay the decision to train. The second is that smaller models are catching larger ones on standard benchmarks, which lifts the floor of base-model capability that ICL exploits. The frozen Llama-3-8B-Instruct used in our experiments is already capable enough to extract signal from book titles at $n = 10$ and to write better recommendations from that signal than a hundred-pair-trained adapter does. Two model generations from now that capability will sit lower on the parameter curve and run faster, which favours ICL further at every $n$ a generative product is likely to see.

The recommendation that comes out of this is sequencing for scoring products and patience for generative ones. A consumer ranking product is best served by starting with ICL, collecting paired preference data from day one, and switching to a trained adapter for users who cross a clear engagement threshold. A consumer generative product should start with ICL, collect the same data, and treat the decision to train as one that requires evidence the gain transfers to the actual output format, not as one that follows automatically from a pairwise-accuracy benchmark. In both cases the data infrastructure is identical, so the choice to invest in collection now does not commit the company to either deployment path later. What changes is which path the empirical record currently supports, and the empirical record, once the format axis is included, supports ICL across a much wider slice of the product landscape than the primary sweep alone suggests.

The open question for the AI community sits exactly here. The methods studied in this thesis sharpen the log-probability margin on the training format and leave the generation distribution largely untouched, which means the gap between offline preference accuracy and deployed personalisation is not just a sampling or evaluation artefact but a substantive property of how small adapters interact with the underlying model. Closing that gap, whether through larger LoRA ranks, training objectives that act on the generation distribution directly, or evaluation protocols that score what the model writes rather than what it ranks, is the next step for work that wants its claims about cold-start alignment to carry over from a paired-completion benchmark to a product. Until that work exists, the practical reading of \cref{sec:results} for most of the LLM economy is that the prompt is still doing the work, and the case for replacing it with a trained adapter needs to be made in the format the user will see.